---
title: |
  | Computational Statistics
  | Lab 3
author: |
  | Group 03
  | Mohsen Pirmoradiyan, Ahmed Alhasan, Yash Pawar
date: "14 Feb 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

## Question 1: Cluster sampling
An opinion pool is assumed to be performed in several locations of Sweden by sending interviewers
to this location. Of course, it is unreasonable from the financial point of view to visit each city. Instead, a decision was done to use random sampling without replacement with the probabilities proportional to the number of inhabitants of the city to select 20 cities. Explore
the file population.xls. Note that names in bold are counties, not cities.

1. Import necessary information to R.
```{r}
RNGversion(min(as.character(getRversion()), "3.6.2"))
set.seed(12345, kind="Mersenne-Twister", normal.kind="Inversion")

##Question 1: Cluster sampling
population <- read.csv2("../Data/population.csv", stringsAsFactors = FALSE)
```

2. Use a uniform random number generator to create a function that selects 1 city from the
whole list by the probability scheme offered above (do not use standard sampling functions
present in R).

3. . Use the function you have created in step 2 as follows:
(a) Apply it to the list of all cities and select one city
(b) Remove this city from the list
(c) Apply this function again to the updated list of the cities
(d) Remove this city from the list
(e) ... and so on until you get exactly 20 cities.
```{r}
sampler <- function(data, n){
  #taking the cumsum for the probabilities means 
  #we put them in a period from 0 to 1 without overlapping
  data$prob <- cumsum(data[,2] / sum(data[,2]))
  
  cities <- as.character()
  pop    <- as.numeric()
  for(i in 1:n){
    rand <- runif(1)
    #applying the generalized inverse distribution function which takes only the nearest   
    #CMD value that is equal or larger than the probability value ..cities with higher population 
    #cover longer periods so they have higher chances to be closest to the roll
    cities[i] <- data[which.min((data$prob-rand)[which((data$prob-rand)>=0)]),][[1]]
    pop[i]    <- data[which(data[,1] == cities[i]),][[2]]
    data      <- data[-which(data[,2] == pop[i]),]
  }
  res <- data.frame(Municipality = as.character(cities), Population = as.numeric(pop))
  return(res)
}
```

4. Run the program. Which cities were selected? What can you say about the size of the
selected cities?
```{r}
sampler(data = population, n = 20)
```
\definecolor{myblue}{HTML}{00007D}
- \textcolor{myblue}{Most of the selected cities are the ones that have larger population (higher probability)}

5. Plot one histogram showing the size of all cities of the country. Plot another histogram
showing the size of the 20 selected cities. Conclusions?
```{r}
par(mfrow=c(1,2))
hist(population$Population, breaks = 50, xlab = "population", main = "All Cities")
hist(sampler(data = population, n = 20)$Population, breaks = 50, xlab = "population", main = "Sample of 20 Cities")
```
- \textcolor{myblue} {Since we are trying to sample from a discrete distribution we obtain the CDF by cumulatively adding up the individual probabilities for the cities and selecting the nearest city that has it's CMD value equal or larger than the probability generated from the uniform distribution of the CDF, based on the generalized inverse distribution function:}

$$ \color{myblue} {F^{-1}(U) = inf\{x \in {\rm I\!R} : F(x) \geq U\}}$$


\newpage
## Question 2: Different distributions
The double exponential (Laplace) distribution is given by formula:
$$DE(\mu, \alpha) = \frac{\alpha}{2} \ \exp(-\alpha|x - \mu|)$$

1. Write a code generating double exponential distribution DE(0,1) from Unif(0,1) by using
the inverse CDF method. Explain how you obtained that code step by step. Generate
10000 random numbers from this distribution, plot the histogram and comment whether
the result looks reasonable.

- \color{myblue} From the PDF of the lablace distribution $DE(\mu, \alpha) = \frac{\alpha}{2} \ \exp(-\alpha|x - \mu|)$ we get the CMD by integrating.

$$\color{myblue} {F(x \geq \mu ) = \int_{-\infty}^{0}\frac{1}{2} \ exp(x) \ dx\ + \int_{0}^{x}\frac{1}{2} \ exp(-x) \ dx}$$ 
$$\color{myblue} {= \frac{1}{2} \ \bigg[exp(x)\bigg]_{-\infty}^0 + \frac{1}{2}\bigg[-exp(-x)\bigg]_0^x}$$
$$\color{myblue} {= 1 - \frac{1}{2}exp(-x)}$$
\newline
$$\color{myblue} {F(x < \mu ) = \int_{-\infty}^{x}\frac{1}{2} \ exp(x) \ dx}$$
$$\color{myblue} {= \frac{1}{2} \ \bigg[exp(x)\bigg]_{-\infty}^x}$$
$$\color{myblue} {= \frac{1}{2}exp(x)}$$
\newline

- \color{myblue} by substituting $F^{-1}F(x)$ in x, where we have $U=F(x)$ we get $F^{-1}(U)$
$$\color{myblue} {U = 1 - \frac{1}{2}exp(-x) \;\;\; \Rightarrow \;\;\; F^{-1}(U) = -ln \ 2(1 - U) \;\;\; ... \;\;\; for \; U\geq \frac{1}{2}}$$
$$\color{myblue} {U = \frac{1}{2}exp(x) \;\;\; \Rightarrow \;\;\; F^{-1}(U) = ln \ 2 U \;\;\; ... \;\;\; for \; U < \frac{1}{2}}$$
\newline 
- \color{myblue} Using the inverse function $F^{-1}(U)$ we can obtain the lablace probability against each uniform probability.
```{r fig.height=3.5}
inverse_cmd <- function(n){
  u <- runif(n)
  
  inverse <- c()
  for(i in 1:n){
    if(u[i] < 0.5){
      inverse[i] <- log(2 * u[i])
    }
    else{
      inverse[i] <- -log(2 * (1 - u[i]))
    }
  }
  return(inverse)
}

x <- inverse_cmd(10000)
hist(x, breaks = 50, main = "Sample of 10000 points from Lablace Distribution")
```
- \color{myblue} The resulted sample has close resemblance to the laplace distribution (double exponential) so it looks reasonable.

2. Use the Acceptance/rejection method with DE(0,1) as a majorizing density to generate
N(0,1) variables. Explain step by step how this was done. How did you choose constant
c in this method? Generate 2000 random numbers N(0,1) using your code and plot the
histogram. Compute the average rejection rate R in the acceptance/rejection procedure.
What is the expected rejection rate ER and how close is it to R? Generate 2000 numbers
from N(0,1) using standard rnorm() procedure, plot the histogram and compare the
obtained two histograms.



\textcolor{myblue} {Target density function:}
$$\color{myblue}{f_Y = \frac{1}{\sqrt{2\pi}}\exp(-x^2/2)}$$

\textcolor{myblue} {Majorizing density function:}
$$\color{myblue}{f_X = \frac{1}{2}\exp(-|x|)}$$

$$\color{myblue}{cf_X \geq f_Y}$$

$$\color{myblue}{c = \frac{2}{\sqrt{2\pi}}\exp(|x|-x^2/2)}$$
\textcolor{myblue} {We set c to 0 to find the value of x}
$$\color{myblue}{\frac{2}{\sqrt{2\pi}}(-x+\frac{x}{|x|})\exp(|x|-x^2/2)=0}$$

\textcolor{myblue} {$\Rightarrow$ for $x\geq 0$  $\Rightarrow$  $x=1$ and for  $x < 0$ $\Rightarrow$ $x=-1$}

$$\color{myblue}{c = \frac{2}{\sqrt{2\pi}}\exp(1-0.5)=\frac{2\sqrt{e}}{\sqrt{2\pi}}}$$
```{r}
c = 2*sqrt(exp(1))/sqrt(2*pi)
cat("The majorizing constant:", c)
```

```{r}
# Target Density Function -- Normal Distribution
fy <- function(x){
  exp(-0.5 * x ^ 2) / sqrt(2 * pi)
}

# Majorizing Density Function -- Laplace Distribution
fx <- function(x){
  exp(-abs(x)) / 2
} 
  
accept_reject <- function(n){
  R <- 0
  Y <- vector(length = n)
  for(i in 1:n){
    repeat {
      y <- inverse_cmd(1)
      U <- runif(1)
      h <- fy(y) / (c * fx(y))
      if(U <= h){
        Y[i] <- y 
        break
        }
      else{R = R+1}
    }
  }
  return(list(Y=Y, Reject=R))
}

set.seed(12345)
Z <- accept_reject(n = 2000)
cat("The number of rejections: ",Z$Reject)

hist(Z$Y, xlab = "Z", main = "Histogram of standard normal distribution", breaks = seq(-4,4,0.1))
```

\textcolor{myblue} {The number of times required to the event "Acceptance" occured is a random variable which has a geometric distribution with success probability of:}

$$\color{myblue}{p = Pr(U\leq \frac{f(Y)}{cg(Y)}|Y=y)}$$


\textcolor{myblue} {Considerind the density distribution $g(Y)$ and the value $\frac{f(y)}{cg(y)}$ we have:}
$$\color{myblue}{p = \int_{-\infty}^{\infty}\frac{f(y)}{cg(y)}g(y)dy=\frac{1}{c}\int_{-\infty}^{\infty}f(y)dy=\frac{1}{c}}$$

```{r}
cat("Expected Rejection Rate: ER = ", 1 - (1/c))
cat("Rejection Rate: R = ", Z$Reject/(Z$Reject+2000))
```
\textcolor{myblue} {The rejection rate "R" is very close to the expected rejection rate "ER"}

```{r}
set.seed(12345)
Z = rnorm(2000)
hist(Z, main = "Histogram of standard normal distribution using R function", breaks = seq(-4,4,0.1))
```
\textcolor{myblue}{The histograms looks fairly similar i.e. generated from normal distribution}

\newpage
## Appendix
```{r eval=FALSE}

```












